---
title: "Supermarket Simulation"
author: "Harindu Perera"
date: "2025-07-14"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Variables that fully describe the status of the system at any point in time:

### System State Variables

    Number of customers currently waiting in each queue.

    Status of each checkout counter (whether busy or free).

    Total number of customers currently in the supermarket

### Counter Variables

    Number of arrivals by time t

    Number of departures after service completion by time t


## Possible Events during the simulation:

Customer arrival : A new customer arrives at the supermarket.

Shopping completion : A customer finishes shopping and moves to the checkout area.

Checkout queue joining event: Customer chooses a checkout counter with the shortest queue and joins it.

Service start event: A cashier becomes free and begins servicing the next customer in their queue.

Service completion event: A customer finishes checkout service and leaves the system.


```{r warning=FALSE}
supermarket_simulation <- function(seed) {
  set.seed(seed)
  
  arrival.time = function(){ 
    
    t.start = 0
    t.end = 13*60 
    lambda = 5
    t.current = t.start
    arrival_times = c()
    
    while (t.current < t.end){
      u = runif(1)
      t.current = t.current + (-log(u))/lambda
      
      if(t.current < t.end) 
        arrival_times = c(arrival_times, t.current)
    }
    return(arrival_times)
  }
  
  Arrivals_in_a_Day = arrival.time()
  no_of_customers = length(Arrivals_in_a_Day)
  
  shopping.times = rlnorm(no_of_customers, meanlog = 0, sdlog = 0.5)
  
  customer.data <- data.frame(
    Customer_No = seq_len(no_of_customers),
    Arrival_Time = Arrivals_in_a_Day,
    Shopping_Time = shopping.times
  )
  
  customer.data$Checkout_Start <- customer.data$Arrival_Time + customer.data$Shopping_Time
  
  service_rates <- c(1/2.5, 1/3, 1/3.5, 1/4)
  
  customer.data$Counter_Assigned <- NA
  customer.data$Service_Start <- NA
  customer.data$Service_End <- NA
  customer.data$Waiting_Time <- NA
  
  # Track the Time when each counter becomes free to serve the next customer
  counter_free_time <- rep(0, 4)
  # Accumulates the total time each counter spends actively serving customers
  counter_busy_time <- rep(0, 4)
  
  # Sort the customer data by their checkout start time to process events in chronological order.
  sorted_customers <- customer.data[order(customer.data$Checkout_Start), ]
  
  for (i in 1:nrow(sorted_customers)) {
    # Get the original index of the current customer from the unsorted data frame
    customer_index <- which(customer.data$Customer_No == sorted_customers$Customer_No[i])
    # Get the time this customer arrives at the checkout area
    checkout_time <- sorted_customers$Checkout_Start[i]
    
    # The current queue length for each counter. A customer is considered in a queue if they are assigned to that counter AND their service has not yet ended by the current 'checkout_time'.
    queue_lengths <- sapply(1:4, function(x) {
      sum(customer.data$Counter_Assigned == x & 
            customer.data$Service_End > checkout_time, na.rm = TRUE)
    })
    
    min_length <- min(queue_lengths)
    # Identify all counters that have this minimum queue length
    shortest_counters <- which(queue_lengths == min_length)
    
    # Choose a counter for the current customer. If there's a tie (multiple shortest queues), randomly sample one of them. Otherwise, choose the single shortest counter.
    chosen_counter <- if(length(shortest_counters) > 1) {
      sample(shortest_counters, 1)
    } else {
      shortest_counters
    }
    
    # Assign the chosen counter to the current customer
    customer.data$Counter_Assigned[customer_index] <- chosen_counter
    
    service_start <- max(checkout_time, counter_free_time[chosen_counter])
    customer.data$Service_Start[customer_index] <- service_start
    
    service_time <- rexp(1, service_rates[chosen_counter])
    service_end <- service_start + service_time
    customer.data$Service_End[customer_index] <- service_end
    
    customer.data$Waiting_Time[customer_index] <- service_start - checkout_time
    
    counter_free_time[chosen_counter] <- service_end
    
    counter_busy_time[chosen_counter] <- counter_busy_time[chosen_counter] + service_time
  }
  
  customer.data$Total_Time <- customer.data$Service_End - customer.data$Arrival_Time
  
  avg_waiting_in_counter <- sapply(1:4, function(i) {
    mean(customer.data$Waiting_Time[customer.data$Counter_Assigned == i], na.rm=TRUE)
  })
  
  avg_total_time <- mean(customer.data$Total_Time)
  
  simulation_end_time <- max(customer.data$Service_End)
  utilization <- counter_busy_time / simulation_end_time
  
  total_customers_served <- nrow(customer.data)
  
  return(list(
    avg_waiting_in_counter = avg_waiting_in_counter,
    avg_total_time = avg_total_time,
    utilization = utilization,
    total_customers_served = total_customers_served,
    shopping_times = shopping.times
    
  ))
}

```

```{r}
n_runs <- 100

results <- lapply(1:n_runs, supermarket_simulation)

avg_wait_matrix <- matrix(NA, n_runs, 4) 
avg_total_vector <- numeric(n_runs)      
util_matrix <- matrix(NA, n_runs, 4)
total_cust_vector <- numeric(n_runs)

for (i in 1:n_runs) {
  avg_wait_matrix[i, ] <- results[[i]]$avg_waiting_in_counter
  avg_total_vector[i] <- results[[i]]$avg_total_time
  util_matrix[i, ] <- results[[i]]$utilization
  total_cust_vector[i] <- results[[i]]$total_customers_served
}

summary_results <- data.frame(
  Counter = 1:4,
  Mean_Avg_Waiting_Time = colMeans(avg_wait_matrix),
  Mean_Utilization = colMeans(util_matrix)
)
```


```{r}
print(round(summary_results, 3))

```

```{r}
cat(sprintf(
  "\nAverage total time in supermarket: %.2f minutes (95%% CI: %.2f - %.2f)\n",
  mean(avg_total_vector),
  quantile(avg_total_vector, 0.025),
  quantile(avg_total_vector, 0.975)
))
cat(sprintf(
  "Average number of customers served per day: %.2f (95%% CI: %.2f - %.2f)\n",
  mean(total_cust_vector),
  quantile(total_cust_vector, 0.025),
  quantile(total_cust_vector, 0.975)
))

```


```{r,warning=FALSE}
library(ggplot2)

all_shopping_times <- unlist(lapply(results, function(x) x$shopping_times))
df_shopping <- data.frame(ShoppingTime = all_shopping_times)

# Histogram for shopping time
ggplot(df_shopping, aes(x = ShoppingTime)) +
  geom_histogram(bins = 30, fill = "yellow", color = "black") +
  ggtitle("Histogram of Shopping Times (All Simulation Runs)") +
  xlab("Shopping Time (minutes)") +
  ylab("Frequency") +
  theme_minimal()

```


```{r,warning=FALSE}
library(ggplot2)
library(reshape2)

df_wait <- as.data.frame(avg_wait_matrix)
names(df_wait) <- paste0("Counter-", 1:4)
df_wait_long <- melt(df_wait, variable.name = "Counter", value.name = "WaitingTime")

# Boxplot for average waiting time
ggplot(df_wait_long, aes(x = Counter, y = WaitingTime, fill = Counter)) +
  geom_boxplot() +
  ggtitle("Distribution of Average Waiting Times by Counter") +
  ylab("Waiting Time (min)") +
  theme_minimal()

# Histogram of average total time in the supermarket per run
hist(avg_total_vector, main = "Avg Total Time in Supermarket per Run",
     xlab = "Time (min)", col = "lightgreen", breaks = 20)

# Boxplot for utilization
boxplot(util_matrix, names = paste("Counter", 1:4),
        main = "Cashier Utilization by Counter",
        ylab = "Utilization", col = "orange")

# Histogram for total customers served per day
df_cust <- data.frame(TotalCustomers = total_cust_vector)
ggplot(df_cust, aes(x = TotalCustomers)) +
  geom_histogram(binwidth = 15, fill = "steelblue", color = "black") +
  ggtitle("Total Customers Served per Day") +
  xlab("Number of Customers")
```



